For Induction-Induction, Induction is Enough

### 2019-11-11 paper/paper.tex
stuff

We work in intensional Martin-Löf type theory, equipped with the following features:
\begin{itemize}
  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when non-dependent.
  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
        when we want to leave the type of the first projection implicit. We use $A
        \times B$ for non-dependent pairs.
  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
    this paper, we use extensional notation, and do not write explicit
    transports over equalities. However, we occasionally indicate by
    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
    $e_1$,\dots,$e_n$.
  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.

@@

We use the naming convention: sorts, operators and equations. Sorts
and operators might have parameters.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \cdot && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\cdot \\
  & {\cdot\eta} && : \{\sigma : \Sub\,\Gamma\,\cdot\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
\end{alignat*}
\todo{We may need more abbrevs here}
Abbreviation:
\[
  \sigma^\uparrow = (\sigma\circ\p,\q)
\]

\todo{add examples where all the different constructors are used}

@@

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, taking the signature of natural numbers:
\[
  \Gamma := \cdot\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

>>>>

We work in intensional Martin-Löf type theory, equipped with the following features:
\begin{itemize}
  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
    non-dependent. We sometimes omit function arguments, by implicitly
    generalising over variables.
  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
        when we want to leave the type of the first projection implicit. We use $A
        \times B$ for non-dependent pairs.
  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
    this paper, we use extensional notation, and do not write explicit
    transports over equalities. However, we occasionally indicate by
    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
    $e_1$,\dots,$e_n$.
  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.

@@

The theory of IIT signatures is given as an initial category with families
(CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
part constitutes the substitution calculus, while the type formers allow
declaring new sorts and operators in a signature.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Universe} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
\end{alignat*}
\subsubsection{Notation and abbreviations}

We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
for inductive parameters, we have $\app$ as a categorical version of function
application. We recover the usual type-theoretic application as follows:
\begin{alignat*}{5}
  & \blank\appm\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & t \appm u := \app\,t\,[\id\,u]
\end{alignat*}
We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
$p$ composed with itself $n$ times. However, for informal examples we instead
use a nameful notation, where we introduce names in context extension and in the
$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,(A : \U)\,\ext\,(a :
\El\,A)$ in a signature for pointed sets.

\subsubsection{Example signatures}

@@

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, taking the signature of natural numbers:
\[
  \Gamma := \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

### 2019-11-11 paper/universaliit.tex
stuff

\todo{Section on metatheoretic notation missing?}

We work in the setting of extensional type theory, although we occasionally indicate through $e_1,\dots,e_n \#
t$ that $t$ is well-typed thanks to the equalities $e_1$,\dots,$e_n$.
% We are also careful to explain where UIP or functional extensionality that is
% available in extensional type theory are used.

@@

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
\todo{András: prettify below maybe}
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inudctive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

@@

  & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\cdot} \\

@@

 := (e_E : \model{E} = \model{\cdot})\times (\model{\delta} = e_E \#

@@

  & \rlap{(2) Sorts and operators} \\[0.5em]

@@

  % $(e_C : \model{E} = \model{\cdot})$, although

@@

@@

%   But to give a precise meaning to the statement that $\wk_n$ (which weakens at
%   the middle of context) preserve the
%   relation, we define (inductively) the notion of semantic telescopes.
%   First, we define the notion of being a prefix:
% \begin{alignat*}{5}
%   & \blank\leq\blank  && : \model\Con\ra \model\Con\ra\Set \\
%   & {\leq}\cdot  && : \model{\Gamma} \leq \model{\Gamma} \\
%   & \blank{\leq}{\ext}\blank  && : \model{\Gamma} \leq \model{\Delta} \ra (A :
%   \model{\Ty}\Delta) \ra \model{\Gamma}\leq\model{\Delta}\model{\ext}\model{A}
% \end{alignat*}
% Now, a telescope on a semantic context $\model{\Gamma}$ is a semantic context
% $\model{\Delta}$ such that $\model{\Gamma}\leq \model{\Delta}$.

% \begin{alignat*}{5}
%   & \Tel\,\model{\Gamma}  && := (\model{\Delta} : \model{\Con})\times (\model{\Gamma}\leq \model{\Delta})
% \end{alignat*}
% Similarly to the untyped case, we define the merging of a context and a telescope by recursion on the telescope:
% \begin{alignat*}{5}
%   & \blank\model{\merge}\blank  && := (\model{\Gamma} : \model{\Con}) \ra
%   (\model{\Delta} : \Tel\,\model\Gamma) \ra \model{\Con} \\
%   & \model\Gamma \merge (\model\Gamma,{\leq}\cdot) && :=  \model\Gamma \\
%   & \model\Gamma \merge (\model\Delta\model{\ext}\model{A}, \Delta_{\leq} {\leq}{\ext} \model{A}) && :=  (\model\Gamma \model\merge \model\Delta)\model\ext \model{A}\\
% \end{alignat*}
% We then define a relevant relation between syntactic telescopes and semantic telescopes.
% Untyped syntactic telescopes are modeled by untyped contexts, because they are
% list of types.
% A syntactic telescope $\Delta$ is then well formed in a context $\Gamma$ if
% $\Gamma\merge\Delta$ is well formed.
% The relation is defined by recursion on the untyped syntactic telescope:
% \begin{alignat*}{5}
%   & \blank\sim\blank  && := ( \Gamma \merge \Delta \vdash) \ra (\model\Gamma\leq\model\Delta)
%   \ra \Set \\
%   & (\wf\Gamma : \Gamma \merge \emptyu \vdash)\sim\model\Delta  && :=
%   \model\Delta =  {\leq}\cdot\\
%   & (\wf\Delta \extw \wf{A} : \Gamma \merge (\Delta \extu A) \vdash)\sim\model{E}  && :=
%   \sum_{\model\Delta} (\wf\Delta\sim\model\Delta)
%   \times
%   \sum_{\model A} (\wf{A}\sim\model{A}) \times
%   \\
%   & && \qquad
%   \model{E} = \model\Delta {\leq}{\ext} \model{A}
% \end{alignat*}

>>>>

% We are also careful to explain where UIP or functional extensionality that is
% available in extensional type theory are used.

@@

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
\todo{András: prettify below maybe}
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inudctive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

@@

  & \emptyw\sim \model{\Gamma} && := \model{\Gamma}=\model{\boldsymbol{\cdot}} \\

@@

 := (e_E : \model{E} = \model{\boldsymbol{\cdot}})\times (\model{\delta} = e_E \#

@@

  & \rlap{(2) Universe} \\[0.5em]

@@

  % $(e_C : \model{E} = \model{\boldsymbol{\cdot}})$, although

@@

  \label{sec:right_uniqueness}

@@

%   But to give a precise meaning to the statement that $\wk_n$ (which weakens at
%   the middle of context) preserve the
%   relation, we define (inductively) the notion of semantic telescopes.
%   First, we define the notion of being a prefix:
% \begin{alignat*}{5}
%   & \blank\leq\blank  && : \model\Con\ra \model\Con\ra\Set \\
%   & {\leq}\boldsymbol{\cdot}  && : \model{\Gamma} \leq \model{\Gamma} \\
%   & \blank{\leq}{\ext}\blank  && : \model{\Gamma} \leq \model{\Delta} \ra (A :
%   \model{\Ty}\Delta) \ra \model{\Gamma}\leq\model{\Delta}\model{\ext}\model{A}
% \end{alignat*}
% Now, a telescope on a semantic context $\model{\Gamma}$ is a semantic context
% $\model{\Delta}$ such that $\model{\Gamma}\leq \model{\Delta}$.

% \begin{alignat*}{5}
%   & \Tel\,\model{\Gamma}  && := (\model{\Delta} : \model{\Con})\times (\model{\Gamma}\leq \model{\Delta})
% \end{alignat*}
% Similarly to the untyped case, we define the merging of a context and a telescope by recursion on the telescope:
% \begin{alignat*}{5}
%   & \blank\model{\merge}\blank  && := (\model{\Gamma} : \model{\Con}) \ra
%   (\model{\Delta} : \Tel\,\model\Gamma) \ra \model{\Con} \\
%   & \model\Gamma \merge (\model\Gamma,{\leq}\boldsymbol{\cdot}) && :=  \model\Gamma \\
%   & \model\Gamma \merge (\model\Delta\model{\ext}\model{A}, \Delta_{\leq} {\leq}{\ext} \model{A}) && :=  (\model\Gamma \model\merge \model\Delta)\model\ext \model{A}\\
% \end{alignat*}
% We then define a relevant relation between syntactic telescopes and semantic telescopes.
% Untyped syntactic telescopes are modeled by untyped contexts, because they are
% list of types.
% A syntactic telescope $\Delta$ is then well formed in a context $\Gamma$ if
% $\Gamma\merge\Delta$ is well formed.
% The relation is defined by recursion on the untyped syntactic telescope:
% \begin{alignat*}{5}
%   & \blank\sim\blank  && := ( \Gamma \merge \Delta \vdash) \ra (\model\Gamma\leq\model\Delta)
%   \ra \Set \\
%   & (\wf\Gamma : \Gamma \merge \emptyu \vdash)\sim\model\Delta  && :=
%   \model\Delta =  {\leq}\boldsymbol{\cdot}\\
%   & (\wf\Delta \extw \wf{A} : \Gamma \merge (\Delta \extu A) \vdash)\sim\model{E}  && :=
%   \sum_{\model\Delta} (\wf\Delta\sim\model\Delta)
%   \times
%   \sum_{\model A} (\wf{A}\sim\model{A}) \times
%   \\
%   & && \qquad
%   \model{E} = \model\Delta {\leq}{\ext} \model{A}
% \end{alignat*}
